---
author: "Karl W Broman"
title: "Analysis of Diversity Outbred mouse data"
date: "`r Sys.Date()`"
output:
    html_document:
        code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE,
                      fig.width=11, fig.height=6)
options(width=110)
set.seed(20200526)
```

Here we consider the data from [Al-Barghouthi et al.
(2021)](https://doi.org/10.1038/s41467-021-23649-0) on Diversity
Outbred mouse data, and compare the use of an exact hidden Markov
model (HMM) to a more approximate one. The data are available at
[zenodo](https://doi.org/10.5281/zenodo.4265417), and analysis scripts
are at [github](https://github.com/basel-maher/DO_project) as well as
archived at [zenodo](https://doi.org/10.5281/zenodo.4718146).

### Load packages and data

We first load [R/qtl2](https://kbroman.org/qtl2) as well
as the [qtl2convert](https://github.com/rqtl/qtl2convert),
[qtl2fst](https://github.com/rqtl/qtl2fst),
[broman](https://github.com/kbroman/broman),
and [here](https://here.r-lib.org) packages.


```{r load_packages}
library(qtl2)
library(qtl2convert)
library(broman)
library(here)
library(qtl2fst)
```

We create a cache directory to store stuff.

```{r cache_directory}
cachedir <- here("DOapp/_cache")
if(!dir.exists(cachedir)) dir.create(cachedir)
```

We download the data from the [zenodo](https://doi.org/10.5281/zenodo.4265417)
and load it; the `.RData` file contains a single object `cross_basic`.

```{r download_data}
url <- "https://zenodo.org/record/4265417/files/cross_basic_cleaned.Rdata"
file <- file.path(cachedir, basename(url))
if(!file.exists(file)) download.file(url, file)
load(file)
```

The data concern `r n_ind(cross_basic)` Diversity Outbred mice
(from generations `r min(cross_basic$cross_info)` &ndash;
`r max(cross_basic$cross_info)`),
including `r sum(cross_basic$is_female)` females and `r sum(!cross_basic$is_female)` males.
They were genotyped with the GigaMUGA array.
There are a total of `r add_commas(tot_mar(cross_basic))` informative markers.

We create a `"genail"` version, to use the general (but approximate)
HMM.

```{r create_genail_version}
genail <- qtl2convert::cross2_do_to_genail8(cross_basic)
```

### Calculate genotype probabilities

We insert pseudomarkers into the marker so that no two
markers/pseudomarkers are > 0.1 cM apart. We use interpolation to find
the corresponding physical positions.

```{r pseudomarker_maps}
gmap <- insert_pseudomarkers(cross_basic$gmap, step=0.1, stepwidth="max")
pmap <- interp_map(gmap, cross_basic$gmap, cross_basic$pmap)
```

```{r save_pseudomarker_maps, echo=FALSE}
saveRDS(gmap, file.path(cachedir, "gmap.rds"))
saveRDS(pmap, file.path(cachedir, "pmap.rds"))
```

Now we do the genome reconstruction, first with the more "exact" HMM.
We assume 0.2% genotyping error and use the [Carter-Falconer map
function](https://www.ias.ac.in/article/fulltext/jgen/050/02/0307-0323).

The results are enormous, so here we will use
[qtl2fst](https://github.com/rqtl/qtl2fst) to save memory, calculating
them but storing them on disk.

```{r calc_genoprob_do, eval=FALSE}
dir <- file.path(cachedir, "probs_do")
if(!dir.exists(dir)) dir.create(dir)

fpr_do <- calc_genoprob_fst(cross_basic, "probs_do", dir,
                            gmap, error_prob=0.002,
                            map_function="c-f", cores=0)
```

```{r calc_genoprob_do_really, echo=FALSE}
dir <- file.path(cachedir, "probs_do")
if(!dir.exists(dir)) dir.create(dir)
file <- file.path(dir, "probs_do_fstindex.rds")
if(file.exists(file)) {
    fpr_do <- readRDS(file)
} else {
<<calc_genoprob_do>>
}
```

We will also convert them to allele dosages, again using the qtl2fst
package to save them on disk.

```{r calc_aprobs_do, eval=FALSE}
dir <- file.path(cachedir, "aprobs_do")
if(!dir.exists(dir)) dir.create(dir)
fapr_do <- genoprob_to_alleleprob_fst(fpr_do, "aprobs_do", dir, cores=0)
```

```{r calc_aprobs_do_really, echo=FALSE}
dir <- file.path(cachedir, "aprobs_do")
if(!dir.exists(dir)) dir.create(dir)
file <- file.path(dir, "aprobs_do_fstindex.rds")
if(file.exists(file)) {
    fapr_do <- readRDS(file)
} else {
<<calc_aprobs_do>>
}
```

We now do the same thing using the `"genail8"` cross type that gives
the approximate HMM. First the 36-state genotype probabilities.

```{r calc_genoprob_genail, eval=FALSE}
dir <- file.path(cachedir, "probs_genail")
if(!dir.exists(dir)) dir.create(dir)

fpr_genail <- calc_genoprob_fst(genail, "probs_genail", dir,
                                gmap, error_prob=0.002,
                                map_function="c-f", cores=0)
```

```{r calc_genoprob_genail_really, echo=FALSE}
dir <- file.path(cachedir, "probs_genail")
if(!dir.exists(dir)) dir.create(dir)
file <- file.path(dir, "probs_genail_fstindex.rds")
if(file.exists(file)) {
    fpr_genail <- readRDS(file)
} else {
<<calc_genoprob_genail>>
}
```

We will also convert them to allele dosages, again using the qtl2fst
package to save them on disk.

```{r calc_aprobs_genail, eval=FALSE}
dir <- file.path(cachedir, "aprobs_genail")
if(!dir.exists(dir)) dir.create(dir)
fapr_genail <- genoprob_to_alleleprob_fst(fpr_genail, "aprobs_genail",
                                          dir, cores=0)
```

```{r calc_aprobs_genail_really, echo=FALSE}
dir <- file.path(cachedir, "aprobs_genail")
if(!dir.exists(dir)) dir.create(dir)
file <- file.path(dir, "aprobs_genail_fstindex.rds")
if(file.exists(file)) {
    fapr_genail <- readRDS(file)
} else {
<<calc_aprobs_genail>>
}
```

Note that when I calculated the genotype probabilities directly
(without using the qtl2fst package) using an 8-core linux laptop with
64 GB RAM, it took
2100 sec with the `"do"` cross type and
1594 sec with the `"genail8"` cross type, so the approximate method
took about 75\% as much time.

## Genome scans for QTL

We'll now perform genome scans to see how the results compare. We'll
use a linear mixed model, with a random polygenic effect to account
for relationships, using the "leave-one-chromosome-out" (LOCO) method.

So first we calculate the kinship matrices.

```{r calc_kinship, eval=FALSE}
k_do <- calc_kinship(fapr_do, "loco", cores=0)
k_genail <- calc_kinship(fapr_genail, "loco", cores=0)
```

```{r calc_kinship_really, eval=TRUE}
kfile_do <- file.path(cachedir, "kinship_loco_do.rds")
kfile_genail <- file.path(cachedir, "kinship_loco_genail.rds")
if(file.exists(kfile_do) && file.exists(kfile_genail)) {
    k_do <- readRDS(kfile_do)
    k_genail <- readRDS(kfile_genail)
} else {
<<calc_kinship>>
    saveRDS(k_do, kfile_do)
    saveRDS(k_genail, kfile_genail)
}
```

We set up some covariates to use; see the
[`map_qtl.R`](https://github.com/basel-maher/DO_project/blob/master/src/map_qtl.R)
file in the [code
repository](https://github.com/basel-maher/DO_project/) for the
[Al-Barghouthi et al.
(2021)](https://doi.org/10.1038/s41467-021-23649-0) paper.

```{r set_up_covariates}
covar <- as.matrix(cross_basic$covar)
covar[,"sex"] <- (covar[,"sex"] == "M")*1
covar <- covar[,!(colnames(covar) %in% c("sac_date", "body_length", "ngen"))]
covar <- apply(covar,2,as.numeric)
rownames(covar) <- rownames(cross_basic$covar)
```

And now we perform the genome scans for a set of selected phenotypes.
For most of them, we take logs; for Wpy we take log(x+1); and there
are a set of 6 that we leave untransformed.

```{r phenotypes}
qcol <- c(TMD="uCT_Ct.TMD",
          Ma.Ar="uCT_Ma.Ar",
          Tt.Ar="uCT_Tt.Ar",
          Ct.Por="uCT_Ct.porosity",
          ML="ML",
          pMOI="uCT_pMOI",
          "Ct.Ar/Tt.Ar"="uCT_Ct.Ar.Tt.Ar",
          Imax="uCT_Imax",
          BMD="uCT_BMD",
          W="bending_work_to_yield",
          Wpy="bending_work_post_yield",
          Dfx="bending_disp_at_frax",
          DFmax="bending_disp_at_max_load",
          Fmax="bending_max_load",
          Ffx="bending_frax_load",
          Ct.Ar="uCT_Ct.Ar",
          Imin="uCT_Imin",
          Ct.Th="uCT_Ct.Th",
          Tb.Sp="uCT_Tb.Sp",
          Tb.N="uCT_Tb.N")

# note that uCT_Ct.porosity has a huge batch effect
# (first three batches have values < 0.5; last 8 have values > 0.5)

# phenotypes to transform (see https://github.com/basel-maher/DO_project/blob/master/src/map_qtl.R#L73-L75)
take_logp1 <- qcol["Wpy"]
take_log <- pheno_names(cross_basic)[c(6:14,16,17,21,23:33,34,35,37:41,43:49,51,52,54:58,61:70,72,74,76)] %win%
                (qcol %wnin% take_logp1)
no_log <- pheno_names(cross_basic)[c(15,18,19,20,22,36,42,50,53,59,60)] %win% qcol

# double-check that I got them all
stopifnot(length(take_logp1) + length(no_log) + length(take_log) == length(qcol))

# pull out and transform the phenotypes
phe <- cross_basic$pheno[,qcol]
phe[,take_log] <- log(phe[,take_log])
phe[,take_logp1] <- log(phe[,take_logp1] + 1)

# change the column names
colnames(phe) <- names(qcol)
```

Now, to perform the genome scans.

```{r scan1, eval=FALSE}
out_scan1_do <- scan1(fapr_do, phe, k_do, addcovar=covar, cores=0)
out_scan1_genail <- scan1(fapr_genail, phe, k_genail, addcovar=covar, cores=0)
```

```{r scan1_really, echo=FALSE}
file1 <- file.path(cachedir, "scan1_do.rds")
file2 <- file.path(cachedir, "scan1_genail.rds")
if(file.exists(file1) && file.exists(file2)) {
    out_scan1_do <- readRDS(file1)
    out_scan1_genail <- readRDS(file2)
} else {
<<scan1>>
    saveRDS(out_scan1_do, file1)
    saveRDS(out_scan1_genail, file2)
}
```


## Comparisons

The genotype probabilities can hardly be distinguished.

```{r probs_diff, eval=FALSE}
p_d <- sapply(names(fpr_do), function(chr) max(abs(fpr_do[[chr]] - fpr_genail[[chr]])))
```

```{r probs_diff_really, echo=FALSE}
file <- file.path(cachedir, "probs_diff.rds")
if(file.exists(file)) {
    p_d <- readRDS(file)
} else {
<<probs_diff>>
    saveRDS(p_d, file)
}
```

The maximum difference was `r myround(max(p_d), 3)` on the
`r names(which.max(p_d))` chromosome followed by a difference of
`r myround(sort(p_d, decreasing=TRUE)[2], 4)` on chromosome
`r names(sort(p_d, decreasing=TRUE))[2]`. For that reason, the QTL
mapping results were hardly different. Across all `r ncol(out_scan1_do)`
traits considered, the maximum difference in LOD scores in the two
sets of results was `r myround(maxlod(abs(out_scan1_do - out_scan1_genail)), 3)`.

Here's a plot of the LOD curves across the genome for tissue mineral
density (TMD). The differences are too small to be distinguished by eye.

```{r plot_tmd_scan}
layout(rbind(1,2), height=c(3,1))
par(mar=c(3.1, 3.1, 1.1, 0.6))
plot(out_scan1_do, pmap, lod="TMD")
plot(out_scan1_genail, pmap, lod="TMD", add=TRUE, col="violetred", lty=3)
plot(out_scan1_do - out_scan1_genail, pmap, ylim=c(-0.02, 0.02), lod="TMD",
     ylab="LOD(DO) - LOD(genAIL8)")
```



## Session info

Here are the details on the version of R and the versions of the R
packages that I used.

```{r session_info}
sessionInfo()
```
